---
title: "Version Control with Git"
#author: "Liming Wang"
#date: "8/21/2017"
output: 
  html_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Version Control with Git

## Install Git {#install-git}

### for windows
1. Download the Git for Windows installer from https://git-scm.com/download/win.
1. Run the installer and follow the steps bellow:
    1. Click on "Next".
    1. Click on "Next".
    1. Keep "**Use Git from the Windows Command Prompt**" selected and click on "Next". If you forgot to do this, rerun the installer and select the appropriate option.
    1. Click on "Next".
    1. Keep "**Checkout Windows-style, commit Unix-style line endings**" selected and click on "Next".
    1. Keep "**Use Windows' default console window**" selected and click on "Next".
    1. Click on "Install".
    1. Click on "Finish".
1. If your "HOME" environment variable is not set (or you don't know what this is):
    1. Open command prompt (Open Start Menu then type 'cmd' (without quotes) and press [Enter])
    1. Type the following line into the command prompt window exactly as shown:
         `setx HOME "%USERPROFILE%"`
    1. Press [Enter], you should see `SUCCESS: Specified value was saved`.
    1. Quit command prompt by typing exit then pressing [Enter]

### For Mac 
Install [git-osx-installer](https://sourceforge.net/projects/git-osx-installer/files/).

### For Linux 
install `git' from your distro's package manager.

## Lesson
Adapted from [Version Control with Git by Software Carpentry](http://swcarpentry.github.io/git-novice/).

- [A quick overview of Version Control](http://swcarpentry.github.io/git-novice/01-basics/)
- Start off by registering a github account at https://github.com
- Use your github user name and email to config git. Open a Shell/Command line window from RStudio menu **Tools/Shell...**, and modify and run the following commands:
```{r, eval=FALSE}
git config --global user.name "Liming Wang"
git config --global user.email "lmwang+datascience@gmail.com"
git config --global color.ui "auto"
```

- Let's turn the RStudio project you have been working on into a git repository and commit it to GitHub by following these steps: http://happygitwithr.com/existing-github-last.html

- [Cache credentials](http://happygitwithr.com/credential-caching.html)
- [Collaborating](http://swcarpentry.github.io/git-novice/08-collab/)
- [Conflicts](http://swcarpentry.github.io/git-novice/09-conflict/)
- [Review of common git commands](http://happygitwithr.com/usage-git-cmds.html)

## Exercise
1. If you haven't yet, commit your RStudio project to github;
1. Add the GitHub account of your team partner as a collaborator to your repo;
1. Exchange URL to your github repo with your team partner;
1. Checkout your team partner's repo in RStudio (File/New Project ..., then Verision Control);
1. Review his/her code and add comments for suggestions where the code can be improved;
1. Commit your change (comments) with a proper commit message;
1. Push your commit to GitHub;
1. After your partner pushes his/her commit, switch back to your own repo & pull to get the updated files with your partner's comments;
1. Responded to the comments with your own comments or changes to your code, commit (with a commit message) and then push to GitHub;
1. Repeat the above steps a few times.

# Resources:
1. [Happy Git with R](http://happygitwithr.com/)
1. [try Git](https://try.github.io)
1. [Pro Git](https://git-scm.com/book/en/v2), a book by Scott Chacon and Ben Straub
